---
title: "VAR_Cholesky"
output: html_document
date: "2025-07-11"
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tseries)
library(quantmod)
library(ggplot2)
library(vars)
library(data.table)
library(jtools)
library(dplyr)
library(zoo)
library(forecast)
library(tidyr)
library(dplyr)
library(tibble)
library(patchwork)
#install.packages("stargazer")
library(stargazer)
library(lpirfs)
library(tidyverse)
library(svars)

```

## Data Loading and Making a Data Frame

We first load the data and the transform it to turn it into a time-series object. We then create a "data" dataframe.

```{r dataframe, include=FALSE}
getwd()
data <- read_excel("../Data/Cholesky.xlsx")

#Transform data
y <- ts(log(data$y)*100, start = c(2000, 1), freq = 12)  # output
p <- ts(log(data$hcpi)*100, start = c(2000, 1), freq = 12)  # consumer inflation
i <- ts(data$i, start = c(2000, 1), freq = 12)
carbon <- ts(data$carbon, start=c(2000,1), freq=12)
u <- ts(data$u, start=c(2000,1), freq=12)

# Plot data
data <- cbind(carbon, p, u,y, i)
data <- as.data.frame(data)
colnames(data) <- c("carbon","p","u", "y", "i")
plot.ts(data)
```

## VAR estimation

We estimate the VAR and estimate the SVAR - sturctural impulse responses with packages.

```{r SVAR_package, include = TRUE, warning=FALSE}
#VAR estimation
var.est1 <- vars::VAR(data, p = 3,type = "cons")

# Setting up the SVAR with Cholesky decomposition
a.mat <- diag(5)
diag(a.mat) <- NA
a.mat[2, 1] <- NA
a.mat[3, 1] <- NA
a.mat[3, 2] <- NA
a.mat[4, 1] <- NA
a.mat[4, 2] <- NA
a.mat[4, 3] <- NA
a.mat[5, 1] <- NA
a.mat[5, 2] <- NA
a.mat[5, 3] <- NA
a.mat[5, 4] <- NA
print(a.mat)

b.mat <- diag(5)
diag(b.mat) <- NA
print(b.mat)

svar.one <- SVAR(var.est1, Amat = a.mat, Bmat=b.mat, max.iter = 10000, 
                 hessian = TRUE)
#v svar ne mors dt identity matrixa za B recimo, pol sam matriko vn vrzs!!!
svar.one$A
svar.one$B

# Compute and store IRFs
par(mfrow = c(4, 1), mar = c(2.2, 2.2, 1, 1), cex = 0.6)
irf_y <- irf(svar.one, response = "y", impulse = "carbon", 
             n.ahead = 36, ortho = TRUE, boot = TRUE)
irf_p <- irf(svar.one, response = "p", impulse = "carbon", 
             n.ahead = 36, ortho = TRUE, boot = TRUE)
irf_i <- irf(svar.one, response = "i", impulse = "carbon", 
             n.ahead = 36, ortho = TRUE, boot = TRUE)
irf_u <- irf(svar.one, response = "u", impulse = "carbon", 
             n.ahead = 36, ortho = TRUE, boot = TRUE)
irf_carbon <- irf(svar.one, response = "carbon", impulse = "carbon", 
                  n.ahead = 36, ortho = TRUE, boot = TRUE)

plot(irf_y)
plot(irf_p)
plot(irf_i)
plot(irf_u)
plot(irf_carbon)

# Compute and store IRFs
two.int.pi <- irf(svar.one, response = "p", impulse = "i", 
                  n.ahead = 36, ortho = TRUE, boot = TRUE)
two.int.y <- irf(svar.one, response = "y", impulse = "i", 
                 n.ahead = 36, ortho = TRUE, boot = TRUE)
two.int.i <- irf(svar.one, response = "i", impulse = "i", 
                 n.ahead = 36, ortho = TRUE, boot = TRUE)
two.int.u <- irf(svar.one, response = "u", impulse = "i", 
                 n.ahead = 36, ortho = TRUE, boot = TRUE)
# Plot each IRF
plot(two.int.pi, main = "Response of p to monetary policy shock")
plot(two.int.y, main = "Response of y to monetary policy shock")
plot(two.int.i, main = "Response of i to monetary policy shock")
plot(two.int.u, main = "Response of u to monetary policy shock")
```

Note that ~ is a shorthand for function(.x). We estimate a VAR algebraically, using OLS. 
First method is the matrix way, whereas the second is done equation by equation. Notice that the two methods are virtually the same. 
Whereas you pull the diagonal elements from the varcov matrix in the first case and look at square roots, you just make up a matrix 
out of the varcov for each individual regression. 
```{r VAR_manually, include = TRUE}
#---------------------- Getting the right dataframes ----------------------------
lagged_data <- data %>%
  mutate(across(everything(),
                .fns = list(
                  lag1 = ~lag(.x, 1),
                  lag2 = ~lag(.x, 2),
                  lag3 = ~lag(.x, 3)
                ),
                .names = "{.col}_{fn}"))
lagged_data$cons <- 1 
clean_frame <- na.omit(lagged_data)
Y <- as.matrix(clean_frame[, c("carbon","p", "u", "y", "i")])
regressors <- grep("_lag", names(clean_frame), value=TRUE)
X <- as.matrix(clean_frame[, c(regressors, "cons")])
Y <- t(Y)
X <- t(X)
#-------------------------------- OLS - matrix ------------------------------------------
B_hat <- Y%*%t(X)%*%solve(X%*%t(X)) #OLS matrix (X'X)^-1 X'Y
U_hat <- Y-B_hat%*%X #residual matrix
T <- ncol(X) #nobs
Sigma_u <- (U_hat %*% t(U_hat))/ (T-nrow(X)) #variance covariance matrix of residuals
invXtX <- solve(X%*% t(X))
vcov_B <- kronecker(Sigma_u, invXtX) #variance covariance matrix of regression coefficient
std_err <- sqrt(diag(vcov_B))
se_matrix <- matrix(std_err, nrow = nrow(B_hat), ncol = ncol(B_hat), byrow=TRUE)
colnames(se_matrix) <- rownames(X)
rownames(se_matrix) <- rownames(B_hat)

#VAR estimation
var.est1 <- vars::VAR(data, p = 3,type = "const")
results <- var.est1

# Extract standard errors for each equation
se_list <- lapply(results$varresult, function(eq_summary) {
  coef_table <- summary(eq_summary)$coefficients
  coef_table[, "Std. Error"]
})
coef <- var.est1$varresult$carbon$coefficients
coef
#-------------------------------- OLS - eq-by-eq ------------------------------------------
carbon_reg <- lm(carbon ~ carbon_lag1 + carbon_lag2+carbon_lag3+p_lag1+p_lag2+p_lag3+u_lag1+u_lag2+u_lag3+y_lag1+y_lag2+y_lag3+i_lag1+i_lag2+i_lag3, data=lagged_data)
p_reg <- lm(p ~ carbon_lag1 + carbon_lag2+carbon_lag3+p_lag1+p_lag2+p_lag3+u_lag1+u_lag2+u_lag3+y_lag1+y_lag2+y_lag3+i_lag1+i_lag2+i_lag3, data=lagged_data)
y_reg <- lm(y ~ carbon_lag1 + carbon_lag2+carbon_lag3+p_lag1+p_lag2+p_lag3+u_lag1+u_lag2+u_lag3+y_lag1+y_lag2+y_lag3+i_lag1+i_lag2+i_lag3, data=lagged_data)
u_reg <- lm(u ~ carbon_lag1 + carbon_lag2+carbon_lag3+p_lag1+p_lag2+p_lag3+u_lag1+u_lag2+u_lag3+y_lag1+y_lag2+y_lag3+i_lag1+i_lag2+i_lag3, data=lagged_data)
i_reg <- lm(i ~ carbon_lag1 + carbon_lag2+carbon_lag3+p_lag1+p_lag2+p_lag3+u_lag1+u_lag2+u_lag3+y_lag1+y_lag2+y_lag3+i_lag1+i_lag2+i_lag3, data=lagged_data)
vcov_carbon_reg <- vcov(carbon_reg)
vcov_p_reg <- vcov(p_reg)
vcov_y_reg <- vcov(y_reg)
vcov_u_reg <- vcov(u_reg)
vcov_i_reg <- vcov(i_reg)
se_carbon_reg <- sqrt(diag(vcov_carbon_reg))
se_p_reg <- sqrt(diag(vcov_p_reg))
se_y_reg <- sqrt(diag(vcov_y_reg))
se_u_reg <- sqrt(diag(vcov_u_reg))
se_i_reg <- sqrt(diag(vcov_i_reg))
```

```{r VAR_shocks, include=TRUE}
#--------------- Manual structural shock from a recursive VAR  -----------------
residuals_var <- residuals(var.est1)
residuals_var
T <- nrow(residuals_var)
vcov <- t(residuals_var)%*%residuals_var/(T-15) #need also small sample adjustment
vcov(var.est1)
P <- chol(vcov) #t(P)=solve(svar.one$A)%*%svar.one$B
P <- t(P) 
P
diagonal_elements <- diag(P)
D <- array(0,dim=c(5,5))
diag(D) <- diagonal_elements
standardized_matrix <- P%*%solve(D)
solve(standardized_matrix) #v tem primeru sta enaka A in B!
svar.one$A
structural_matrix <- solve(svar.one$A)%*%svar.one$B
structural_matrix
#----------------------- STRUCTURAL SHOCKS ---------------------------
structural_shocks <- solve(structural_matrix)%*%t(residuals_var)
#oziroma
structural_shocks <- solve(P)%*%t(residuals_var)
structural_shocks <- as.data.frame(t(structural_shocks))
plot(structural_shocks$carbon)
plot(structural_shocks$i)
carbon_shock <- ts(structural_shocks$carbon, start=c(2000,3), frequency=12)
monetary_shock <- ts(structural_shocks$i, start=c(2000,3), frequency=12)
plot(carbon_shock)
carbon_plot <- autoplot(carbon_shock, color="blue") + theme_bw()+ theme(panel.grid = element_blank(),  
                                                        axis.text = element_text(color = "black", size =13),   plot.title = element_text(hjust = 0.5, size = 15),
                                                        axis.title.x = element_text(size = 15),   panel.border = element_rect(color = "black", linewidth = 1)) + ggtitle("Carbon policy shock") + ylab("") + xlab("Year")
monetary_plot <- autoplot(monetary_shock, color="blue") + theme_bw()+ theme(panel.grid = element_blank(),  
                                                              axis.text = element_text(color = "black", size =13),   plot.title = element_text(hjust = 0.5, size = 15),
                                                              axis.title.x = element_text(size = 15),   panel.border = element_rect(color = "black", linewidth = 1)) + ggtitle("Monetary policy shock")+ ylab("")+ xlab("Year")
shock_plot <- carbon_plot + monetary_plot
shock_plot
plot(shock_plot)
ggsave("../Output/shock_plot.pdf", plot = shock_plot, width = 10, height = 3, dpi = 600)
#-------------------------- IRFs ročno -------------------------------
irf <- Phi(var.est1,nstep=39)
nstep <- abs(as.integer(39)) 
  K <- 5 #določi koliko maš y
  p <- 3 #določi koliko je endogenih lagov
  A <- as.array(Acoef(var.est1)) #vsi ocenjeni koeficienti v A matrikah
  if(nstep >= p){
    As <- array(0, dim = c(K, K, nstep + 1))
    for(i in (p + 1):(nstep + 1)){ #create nstep matrices filled with zeros
      As[, , i] <- matrix(0, nrow = K, ncol = K)
    }
  } else {
    As <- array(0, dim = c(K, K, p))
  }
  for(i in 1:p){
    As[, , i] <- A[[i]] #fill the first p matrices with actual coefficients
  }  
  Phi <- array(0, dim=c(K, K, nstep + 1))
  Phi[, ,1] <- diag(K)
  Phi[, , 2] <- Phi[, , 1] %*% As[, , 1]
  if (nstep > 1) {
    for (i in 3:(nstep + 1)) {
      tmp1 <- Phi[, , 1] %*% As[, , i-1]
      tmp2 <- matrix(0, nrow = K, ncol = K)
      idx <- (i - 2):1
      for (j in 1:(i - 2)) {
        tmp2 <- tmp2 + Phi[, , j+1] %*% As[, , idx[j]]
      }
      Phi[, , i] <- tmp1 + tmp2
    }
  }
  return(Phi)
}

#----------------------------- Structural impulse responses z
rownames(irf) <- c("carbon", "p", "u", "y", "i")
for(i in 1: dim(irf)[3]){
  irf[, , i] <- irf[, , i] %*% P #this is how you get one unit shock impulse responses! - one unit shock to y!!! in our case carbon shock
}
irf_carbon = array(0, dim=c(5,40))
for(i in 1:dim(irf)[3]){
  irf_carbon[,i] <- irf[1:5,1,i]
}
# Loop through each row and plot
for (i in 1:nrow(irf_carbon)) {  # Corrected syntax
  plot(irf_carbon[i, ], type = "o", col = "blue", main = paste("Plot for Row", i), xlab = "Index", ylab = "Value")
}
irf
irf_carbon <- as.data.frame(irf_carbon)
rownames(irf_carbon) <- c("Carbon", "Inflation", "Unemployment", "Output", "Interest rate")
colnames(irf_carbon) <- 1:40
irf_carbon <- t(irf_carbon)
irf_carbon <- as.data.frame(irf_carbon)
Response_carbon <- irf_carbon[,1]
Response_carbon <- as.data.frame(Response_carbon)
Response_carbon$Period <- 1:40 
plot <- ggplot(Response_carbon, aes(x = Period, y = Response_carbon)) + 
  geom_line() +
  theme_bw() + 
  labs(title = "Impulse Response of Carbon", x = "Period", y="") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10, color="black"),
    panel.grid=element_blank()
  )
plot
irf
```

